Web-based Supplementary Materials for Combinatorial Mixture Models for Single-Cell Assays with Application to Vaccine Studies by Greg Finak, Steve De Rosa, Mario Roederer and Raphael Gottardo

### 2012-07-31 WebSupplementaryMaterialMIMOSA.tex
Fixed a lot of notations and equations.

\subsection{Constrained beta--binomial model}

@@

\subsection{Computational details for the Dirichlet-multinomial model}
\noindent\textbf{Marginal likelihood derivations}\\
Because our Dirichlet-multinomial is a direct extension of the beta-binomial model, the marginal likelihoods are obtained in the exact same fashion. For a non--responder, the null marginal likelihood for a given subject is:

\[
\begin{split}
Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u)=&\int_0^1Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u,\mathbf{p}_u)\pi(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
=&\int_0^1Pr(\mathbf{n}_s|\boldsymbol{\alpha}_u,\mathbf{p}_u)Pr(\mathbf{n}_u|\boldsymbol{\alpha}_u,\mathbf{p}_u)\pi(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
=&\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_u)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
=&\int_0^1 \frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\prod_k p^{n_{sk}}_{uk} \frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\prod_k p^{n_{uk}}_{uk}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\prod_k p_{uk}^{\alpha_{uk}-1}\mathrm{d}\mathbf{p}_u\\
=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\int_0^1 \prod_k p^{n_{sk}+n_{uk}+\alpha_{uk}-1}_{uk}\mathrm{d}\mathbf{p}_u\\
=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{\mathrm{B}(\mathbf{n}_s+\mathbf{n}_u+\boldsymbol{\alpha}_u)}{\mathrm{B}(\boldsymbol{\alpha}_u)}
\end{split}
\]
%The product over all $P$ observations gives the marginal likelihood,
%\[
%\mathrm{L_0}(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u)=\prod_{i=0}^P \frac{\mathbf{N}_{si}!}{\prod_k \mathbf{n}_{sik}!}\frac{\mathbf{N}_{ui}!}{\prod_k \mathbf{n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{si}+\mathbf{n}_{ui}+\boldsymbol{\alpha}_{u})}{\mathrm{B}(\boldsymbol{\alpha}_{u})}
%\]

Similarly, the alternative marginal likelihood for a given subject is:
\[
\begin{split}
Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s)=&\int_0^1\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_s)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{Dir}(\boldsymbol{\alpha}_s)\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}\int_0^1\prod_k p^{n_{uk}+\alpha_{uk}-1}_{uk}\int_0^1 \prod_k p^{n_{sk}+\alpha_{sk}-1}_{sk}\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{\mathrm{B}(\mathbf{n}_u+\boldsymbol{\alpha}_u)\mathrm{B}(\mathbf{n}_s+\boldsymbol{\alpha}_s)}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}
\end{split}
\]
with marginal likelihood:
\[
\mathrm{L_1}(\boldsymbol{\alpha}_s,\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u)=\prod_{i=0}^P\frac{\mathbf{N}_{si}!}{\prod_k \mathbf{n}_{sik}!}\frac{\mathbf{N}_{ui}!}{\prod_k \mathbf{n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{ui}+\boldsymbol{\alpha}_u)\mathrm{B}(\mathbf{n}_{si}+\boldsymbol{\alpha}_s)}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}
\]
\noindent\textbf{MCMC algorithm}\\
The MCMC algorithm for the dirichlet--multinomial model is analogous to the beta--binomial, above. The parameter vectors $\boldsymbol{\alpha}_s,\boldsymbol{\alpha}_u$ are updated component--wise:
\begin{enumerate}
\item  Update each $\alpha_{sk}$, $\alpha_{uk}$ using a Gaussian symmetric proposal distribution with the variance of each proposal tuned using the approach of~\cite{Gelman:2004tc}. 
\item Update $w$ by Gibbs sampling using the full conditional,
\[
(w|\cdots)\sim\mathrm{Beta}(\sum_i z_i,\sum_i(1-z_i))
\]
\item For each $i$, update $z_i$ using the full conditional,
\[
(z_i|\cdots)\sim \mathrm{B}(1,p_i)
\]
where 
\[
p_i = \frac{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}_{u},\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u)}{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}_{u},\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u)+(1-w)\cdot\mathrm{L_0}(\boldsymbol{\alpha}_{u},|\mathbf{n}_s,\mathbf{n}_u)}
\]
For each parameter component updated in step 1 above, compute the acceptance ratio (e.g. $\alpha_{uk}$, below):
\[
\frac{\mathrm{L}(\alpha^{\text{new}}_{uk},\boldsymbol{\theta}_u,\boldsymbol{\alpha}_s|\cdots)\pi(\alpha^{\text{new}}_{uk})}{\mathrm{L}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\cdots)\pi(\alpha_{uk})}\\
\]

where $\pi$ is the prior distribution of the parameter, and $\boldsymbol{\theta}_u = \{\alpha_{uj} : (j \in \mathbb{Z}) (1\leq j \leq K) (j\neq k)\}$. Again, we have used the same exponential prior with mean 1,000 for each parameter. $\mathrm{L}$ is the complete marginal likelihood conditional on $\mathbf{z}$, defined as
\[
\mathrm{L}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u,\mathbf{z}) = \prod_i \mathrm{L}_{z_i}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u)
\]

@@

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto,node distance=0cm]
%\node at (0,0) (A) {
%\begin{tikzpicture}
%    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
%    \begin{scope} [x={(a.south east)},y={(a.north west)}]
%        \node at (0,1) [font=\small\sffamily] {A} ;
%    \end{scope}
%    \end{tikzpicture}
%    };
%    \node [right=of A] (B) {
%    \begin{tikzpicture}
%    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
%        \begin{scope} [x={(b.south east)},y={(b.north west)}]
%        \node at (0,1) [font=\small\sffamily] {B} ;
%        \end{scope}
%    \end{tikzpicture}
%    };
    \node at (0,0) (A) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
         \begin{scope} [x={(c.south east)},y={(c.north west)}]
                 \node at (0,1) [font=\small\sffamily] {A} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [right=of A] (B) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
        \begin{scope} [x={(d.south east)},y={(d.north west)}]
            \node at (0,1) [font=\small\sffamily] {B} ;
\end{scope}
    \end{tikzpicture}
    };
  \node [below=of A] (C) {
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
    \begin{scope} [x={(e.south east)},y={(e.north west)}]
        \node at (0,1) [font=\small\sffamily] {C} ;
    \end{scope}
    \end{tikzpicture}
    };
    \node [right=of C] (D) {
    \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
        \begin{scope} [x={(f.south east)},y={(f.north west)}]
        \node at (0,1) [font=\small\sffamily] {D} ;
        \end{scope}
    \end{tikzpicture}
    };
    \node [below=of C] (E) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
         \begin{scope} [x={(g.south east)},y={(g.north west)}]
                 \node at (0,1) [font=\small\sffamily] {E} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [right=of E] (F) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
        \begin{scope} [x={(h.south east)},y={(h.north west)}]
            \node at (0,1) [font=\small\sffamily] {F} ;
\end{scope}
    \end{tikzpicture}
    };
 \end{tikzpicture}
   \caption{Unconstrained MIMOSA model fit to data from a model violating model assumptions and to two--sided data with small counts. Data was simulated from a model where proportions were sampled from a truncated normal distribution over $[0,1]$ rather than a Beta distribution.  A) The average ROC from 10 simulation with N=5,000 events. B) The average observed and nominal FDR from 10 simulations with N=5,000 events. C) Average ROC for N=1,000 events D) Average observed and nominal FDR for N=1,000 events. MIMOSA fit to two--sided data were simulated from the standard model with N=1,000 events. E) Average ROC curves from 10 simulations and F) average observed vs nominal FDR from 10 simulations.}
   \label{webfig:simulations_trunc}
\end{figure}
\bibliographystyle{biom}
\bibliography{MIMOSA}

>>>>

\subsection{Constrained beta-binomial model}

@@

\subsection{Computational details for the Dirichlet-multinomial model}
\noindent\textbf{Marginal likelihood derivations}\\
Because our Dirichlet-multinomial is a direct extension of the beta-binomial model, the marginal likelihoods are obtained in the exact same fashion. 
For a given subject, the null marginal likelihood is defined as
\[
\begin{split}
\mathrm{L_0}(\boldsymbol{\alpha}_u|\mathbf{n}_{si},\mathbf{n}_{ui})=&\idotsint\mathrm{Pr}(\mathbf{n}_{si},\mathbf{n}_{ui}|\mathbf{p}_u)\pi(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
=&\idotsint \frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\prod_k p^{n_{sik}}_{uk} \frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\prod_k p^{n_{uik}}_{uk}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\prod_k p_{uk}^{\alpha_{uk}-1}\mathrm{d}\mathbf{p}_u\\
=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\int_0^1 \prod_k p^{n_{sik}+n_{uik}+\alpha_{uk}-1}_{uk}\mathrm{d}\mathbf{p}_u\\
=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{si}+\mathbf{n}_u+\boldsymbol{\alpha}_u)}{\mathrm{B}(\boldsymbol{\alpha}_u)}
\end{split}
\]
%The product over all $P$ observations gives the marginal likelihood,
%\[
%\mathrm{L_0}(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u)=\prod_{i=0}^P \frac{\mathbf{N}_{si}!}{\prod_k \mathbf{n}_{sik}!}\frac{\mathbf{N}_{ui}!}{\prod_k \mathbf{n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{si}+\mathbf{n}_{ui}+\boldsymbol{\alpha}_{u})}{\mathrm{B}(\boldsymbol{\alpha}_{u})}
%\]

\noindent Similarly, the alternative marginal likelihood for a given subject is:
\[
\begin{split}
\mathrm{L_1}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_{ui},\mathbf{n}_{si})
%=&\int_0^1\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_s)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{Dir}(\boldsymbol{\alpha}_s)\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}\idotsint\prod_k p^{n_{uik}+\alpha_{uk}-1}_{uk}\idotsint \prod_k p^{n_{sik}+\alpha_{sk}-1}_{sk}\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{ui}+\boldsymbol{\alpha}_u)\mathrm{B}(\mathbf{n}_{si}+\boldsymbol{\alpha}_s)}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}.
\end{split}
\]
\noindent\textbf{MCMC algorithm}\\
The MCMC algorithm for the Dirichlet-multinomial model is analogous to the beta-binomial, above. The parameter vectors $\boldsymbol{\alpha}_s,\boldsymbol{\alpha}_u$ are updated component-wise:
\begin{enumerate}
\item  Update each $\alpha_{sk}$, $\alpha_{uk}$ using a Gaussian symmetric proposal distribution with the variance of each proposal tuned using the approach of~\cite{Gelman:2004tc}. 
\item Update $w$ by Gibbs sampling using the full conditional,
\[
(w|\cdots)\sim\mathrm{Beta}(\sum_i z_i,\sum_i(1-z_i))
\]
\item For each $i$, update $z_i$ using the full conditional,
\[
(z_i|\cdots)\sim \mathrm{B}(1,p_i)
\]
where 
\[
p_i = \frac{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}_{u},\boldsymbol{\alpha}_s|\mathbf{n}_{si},\mathbf{n}_{ui})}{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}_{u},\boldsymbol{\alpha}_s|\mathbf{n}_{si},\mathbf{n}_{ui})+(1-w)\cdot\mathrm{L_0}(\boldsymbol{\alpha}_{u}|\mathbf{n}_{si},\mathbf{n}_{ui})}
\]
For each parameter component updated in step 1 above, compute the acceptance ratio (e.g. $\alpha_{uk}$, below):
\[
\frac{\mathrm{L}(\alpha^{\text{new}}_{uk},\boldsymbol{\alpha}_{u\{-k\}},\boldsymbol{\alpha}_s|\cdots)\pi(\alpha^{\text{new}}_{uk})}{\mathrm{L}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\cdots)\pi(\alpha_{uk})}\\
\]

where $\pi$ is the prior distribution of the parameter, and $\boldsymbol{\alpha}_{u\{-k\}} = \{\alpha_{uj} :  j\neq k\}$. Again, we have used the same exponential prior with mean 1,000 for each parameter. $\mathrm{L}$ is the complete marginal likelihood conditional on $\mathbf{z}$, defined as
\[
\mathrm{L}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_{si},\mathbf{n}_{ui},\mathbf{z}) = \prod_i \mathrm{L}_{z_i}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_{si},\mathbf{n}_{ui}).
\]

@@

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto,node distance=0cm]
%\node at (0,0) (A) {
%\begin{tikzpicture}
%    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
%    \begin{scope} [x={(a.south east)},y={(a.north west)}]
%        \node at (0,1) [font=\small\sffamily] {A} ;
%    \end{scope}
%    \end{tikzpicture}
%    };
%    \node [right=of A] (B) {
%    \begin{tikzpicture}
%    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
%        \begin{scope} [x={(b.south east)},y={(b.north west)}]
%        \node at (0,1) [font=\small\sffamily] {B} ;
%        \end{scope}
%    \end{tikzpicture}
%    };
    \node at (0,0) (A) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
         \begin{scope} [x={(c.south east)},y={(c.north west)}]
                 \node at (0,1) [font=\small\sffamily] {A} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [right=of A] (B) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
        \begin{scope} [x={(d.south east)},y={(d.north west)}]
            \node at (0,1) [font=\small\sffamily] {B} ;
\end{scope}
    \end{tikzpicture}
    };
  \node [below=of A] (C) {
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
    \begin{scope} [x={(e.south east)},y={(e.north west)}]
        \node at (0,1) [font=\small\sffamily] {C} ;
    \end{scope}
    \end{tikzpicture}
    };
    \node [right=of C] (D) {
    \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
        \begin{scope} [x={(f.south east)},y={(f.north west)}]
        \node at (0,1) [font=\small\sffamily] {D} ;
        \end{scope}
    \end{tikzpicture}
    };
    \node [below=of C] (E) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
         \begin{scope} [x={(g.south east)},y={(g.north west)}]
                 \node at (0,1) [font=\small\sffamily] {E} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [right=of E] (F) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
        \begin{scope} [x={(h.south east)},y={(h.north west)}]
            \node at (0,1) [font=\small\sffamily] {F} ;
\end{scope}
    \end{tikzpicture}
    };
 \end{tikzpicture}
   \caption{Unconstrained MIMOSA model fit to data from a model violating model assumptions and to two-sided data with small counts. Data was simulated from a model where proportions were sampled from a truncated normal distribution over $[0,1]$ rather than a Beta distribution.  A) The average ROC from 10 simulation with N=5,000 events. B) The average observed and nominal FDR from 10 simulations with N=5,000 events. C) Average ROC for N=1,000 events D) Average observed and nominal FDR for N=1,000 events. MIMOSA fit to two-sided data were simulated from the standard model with N=1,000 events. E) Average ROC curves from 10 simulations and F) average observed vs nominal FDR from 10 simulations.}
   \label{webfig:simulations_trunc}
\end{figure}
\bibliographystyle{biom}
\bibliography{MIMOSA}
